---
title: Comparing Firearm Mortality to Firearm Ownership by State
authors:
  - name: Grace Rinehart
    affiliation: Colorado State University
    roles: writing
    corresponding: true
  - name: Kendall Landwehr
    affiliation: Colorado State University
    roles: research
    corresponding: false
  - name: Andrew Zimbelman
    affiliation: Colorado State University
    roles: data collection & analysis
    corresponding: false
bibliography: references.bib
csl: apa.csl
editor: 
  markdown: 
    wrap: 72
excecute:
  echo: true
---

# Introduction, Background & Motivation

Through this study, we hope to address and provide increased analysis of
the problem of firearm mortality in the United States. Gun violence is a
major safety, social, and public health problem within the United States
and requires additional investigation. The current literature suggests
that within developed nations and other high income countries, the U.S.
is an outlier in its rates of gun mortality. Initially linked to
America’s frontier history and culture of citizen-owned guns, the U.S.
continues to protect this 2nd amendment right and tradition of freedom,
protection, and spirit of rebellion (@fortune2018guns). While a
constitutional right, it may also be important to consider the needs of
present-day society and how this right can be protected, yet regulated
for a safer future. The large proportion of gun violence, mass
shootings, and firearm mortality can partially be attributed to the
weapon instrumentality effect. This is the hypothesis that the presence
of guns leads to higher levels of violence, aggression, and lethal
action when faced with a stressful or conflictual situation
(@spieller2020gun). While some argue that aggression and violence will
occur regardless of the weapon used, the capacity for lethality also has
to be considered. It is much easier to unintentionally harm someone with
a firearm than a knife, for instance. Understanding the relationship
between firearm ownership and firearm mortality through this study will
contribute to this system of thinking. If the hypothesis that firearm
ownership and firearm mortality are positively correlated, this study
can serve as further evidence that additional firearm restrictions
should be implemented to protect everyday American people. 

Within the U.S., gun mortality continues to be a major problem for its
population, contributing to both mortality and morbidity. In 2020 alone,
the Centers for Disease Control reported 45,000 firearm-related deaths
with a rate of 6.1 deaths per 100,000 people. This is the highest
reported rate since 1994 with 40% of deaths due to homicide and more
than half due to suicide (@fontanarosa2022unrelenting). Access to
firearms is also shifting suicide rates and the proportion of suicide
deaths by firearm. Within the 45,222 firearm deaths in 2020, 60% of
those were due to suicide. Similarly, among the 45,979 suicide deaths in
the same year, 51% of those were from a firearm injury
(@fontanarosa2022unrelenting). These statistics present a clear need for
additional firearm regulation. Our study hopes to provide further
evidence by illustrating a correlation between firearm ownership and
firearm mortality from the Centers for Disease Control and Prevention
and the World Population Review. 

This study will use one dataset from the Centers for Disease Control and
Prevention
([data](https://www.cdc.gov/nchs/pressroom/sosmap/firearm_mortality/firearm.htm))
and one from the World Population Review
([data](https://worldpopulationreview.com/state-rankings/gun-ownership-by-state#sources)).
The CDC dataset provides information on firearm mortality by state and
this study will specifically focus on data from the year 2022. The
firearm mortality data includes the number of deaths and death rate per
state. More specifically, the death rate is the number of deaths per
100,000 total population. The data also ranks each of the states
according to their mortality rate. Additionally, the CDC provides a
visual representation of the state ranking displayed on a map of the
U.S. using a gradient. 

The dataset used for firearm ownership will be provided by the World
Population Review and is originally from the Pew Research Center. This
source provides information on gun ownership as a percentage of
households that own guns within each state. Additionally, data was
collected on gun ownership licenses. It should be noted that the varying
gun regulations on gun ownership throughout the country make it
difficult to obtain exact numbers for firearm ownership, however, the
Pew Research Center has compiled accurate and reliable data given this
limitation. It is also important to note that while one state may have
more guns, they may also have a higher population. Because of this, it
is important to make the distinction that data is based on the total
percentage of firearm ownership in relation to each state’s population.
Based on the data, some general trends can be noted. Gun ownership is
highest in the South with about 36% of residents owning a gun. The West
and Midwest each have ownership rates of around 30%, and the Northeast
has the lowest rates of gun ownership with about 16% of residents in the
region owning a firearm. Moreover, individuals in rural communities have
the highest rates of firearm ownership and men are more likely to own a
firearm than women. Similar to the firearm mortality data, this source
also provides a visual representation of its data, organized on a map of
the United States along a gradient according to ownership percentage. 

Regarding the methods for this study, our data will first be cleaned and
organized so it is more compatible with the functions within R studio.
This study will narrow its scope to the year 2022 and focus specifically
on the top five states with the highest percentage of firearm ownership.
From here we will join the two data sets by state and conduct multiple
statistical tests to identify the relationship between the two
variables. More specifically, we will use the Shapiro test to test for
normality, followed by the Spearman or Pearson test to analyze
correlation. A linear regression model will also be used. After the
necessary statistics have been run, we will develop ways to visualize
and display the correlation using the ggplot package.

# Data Exploration

Two datasets were used in the research project, which were both obtained
by the Centers for Disease Control & Prevention and the World Population
Review. Our data from the Centers for Disease Control & Prevention
provides us with information related to deaths per state and mortality
rate. The World Population Review gives us information related to gun
ownership that we can use to analyze how the gun ownership percentage in
homes per state impacts the mortality rate. A full join was used in our
data-cleaning process to allow analysis to be done between the data in
the different datasets. The data sets were cleaned beforehand in Excel.
Both data sets were manipulated to include a year (2022) and state
column with matching variable names (AK changed to Alaska, etc).
Information not needed for our analysis, like the URLs column in the gun
mortality data set, was removed for clarity. Data was also cleaned in R
Studio, using read.csv tools. The data was imported into R where a full
join was used to turn our semi-cleaned data into one dataset, renaming
tools were used to make column category names less confusing and
shorter. Filtering tools were used to ensure that only the 50 states are
included in the dataset, such as the District of Columbia, which is
already in the dataset from Washington. From our full join, some data
didn’t carry over, which resulted in the manual importing of data points
into the dataframe. Finally, the data was arranged based on highest to
lowest gun ownership per household percentage, then a head command was
used to filter the data to 5 states. 

```{r}
library(readr)
library(flextable)
library(dplyr)
library(ggplot2)
library(tidymodels)
library(ggpubr)

gun_mortality_2022 <- read.csv("data/gun_mortality_2022.csv")
gun_ownership_2022 <- read.csv("data/gun_ownership_2022.csv")

gun_data <- full_join(gun_mortality_2022, gun_ownership_2022, by = 'STATE')
gun_data <- gun_data |>
  rename(c(Year = YEAR.x,
         State = STATE,
         Death_Rate = RATE, 
         Deaths = DEATHS,
         Household_pct_ownership = GunOwnership_PercentageOfHouseholdsThatOwnGuns_pct_2022,
         GunOwnership = GunOwnership_NumOfGunLicenses_num_2022))|>
        select(State,Death_Rate,Deaths,Household_pct_ownership,GunOwnership,Year)
gun_data <- gun_data |>
  filter (State != "District of Columbia")

# Importing all values that didn't transfer over from the join
gun_data[51,3] <- as.integer(797)
gun_data[51,2] <- as.numeric(19.8)
gun_data[52,3] <- as.integer(830)
gun_data[52,2] <- as.numeric(14.0)
gun_data[53,3] <- as.integer(1316)
gun_data[53,2] <- as.numeric(14.9)
gun_data[36,4] <- as.numeric(54.7)
gun_data[36,5] <- as.integer(103368)
gun_data[46,4] <- as.numeric(44.6)
gun_data[46,5] <- as.integer(423707)
gun_data[49,4] <- as.numeric(45.3)
gun_data[49,5] <- as.integer(95430)
gun_data[51,6] <- as.integer(2022)
gun_data[52,6] <- as.integer(2022)
gun_data[53,6] <- as.integer(2022)

gun_data <- gun_data|>
  arrange(-Household_pct_ownership)

gun_data_con <- gun_data |>
  arrange(-Household_pct_ownership) |>
  head(n = 5)
```

For visualizations and tests, a ggbarplot was first created to visualize
the relationship of interest between household gun ownership and death
rates, This relationship was measured by grouping household gun
ownership % by state, with the color of the ggbarplot being the death
rate in states. This visualization showed that the top 5 states had a
similar death rate as well as gun ownership rate.

```{r}
plot1<- gun_data_con |>
ggbarplot(x = 'State', y = 'Household_pct_ownership',
          fill = "Death_Rate") 
print(plot1)
```

A Shapiro test was used to determine if there was normal distribution
within our desired data. It showed that the null hypothesis was not
rejected which tells us that there is a normal distribution within our
data that will be used for future analysis.

```{r}
shapiro.test(gun_data_con$Household_pct_ownership)

shapiro.test(gun_data_con$Death_Rate)
```

Based on this, a linear regression was used to test for a relationship
between our desired variables. This test rejected the null hypothesis,
which suggests that there is a statistically significant relationship
between gun ownership % per household and death rate. The R-squared
value was in the 80% range, which tells us that there is a strong
relationship between our variables. With this, a scatter plot was made
with a best-fit line highlighting the residuals within the data relative
to the best-fit line. Data may be more strongly correlated, but due to
the sample size, it appears as being less correlated. 

```{r}
plot2 <- ggplot(gun_data_con, aes(x = Household_pct_ownership, y = Death_Rate)) +
  geom_point() +
  geom_smooth(method = lm, color = 'red') +
  labs(x = "Gun % Per Household",
       y = "Death Rate",
       title = "Relationship Between gun % Per Household and Death Rate ") +
  theme_classic()
print(plot2)

linear_mod <- lm(Household_pct_ownership ~ Death_Rate, data = gun_data_con)
summary(linear_mod)
```

# Methods

We have decided to scale back our analysis to look at the top 5 states
with the most gun licenses and compare the gun mortality in those
states. Once our data has been tested for normality using the Shapiro
test, we’ll look at the correlation between the variables using the
Spearman or Pearson tests, depending on whether or not the data is
normally distributed. Once our data has been tested, we’ll analyze our
hypothesis using a linear regression model. We could bring in
supplemental data involving parameters like state population and income
as control variables for our model. One potential challenge in the data
is choosing which variables to use for our analysis. The CDC gun
mortality data set includes both the death rate as a percentage and the
total number of deaths for each state. The World Population Review data
set for gun ownership includes the percentage of households that own
guns and gun ownership licenses in each state. Choosing the correct
variable for our analysis is important to prevent our data from skewing,
hence why adding a state population data set may be beneficial.

It would be best for us to conduct our analysis using per capita values,
as those will give us more accurate representations of each state's gun
mortality and gun ownership rates. Some challenges in the methods would
be the lack of in-depth analysis. Due to our variables expected strong,
positive correlation, there is only so much analysis we can do. This is
another reason why adding another data set could prove to be useful. Our
current selected methods will allow us to examine out hypothesis in the
most direct manner, and our correlations will be easy to see in graphs
and figures with a linear regression model. The methods we have outlined
provide the most streamline analysis, and will allow us to reach our
goal of proving our hypothesis efficiently.
