---
title: Comparing Firearm Mortality to Firearm Ownership by State
authors:
  - name: Grace Rinehart
    affiliation: Colorado State University
    roles: writing
    corresponding: true
  - name: Kendall Landwehr
    affiliation: Colorado State University
    roles: research
    corresponding: false
  - name: Andrew Zimbelman
    affiliation: Colorado State University
    roles: data collection & analysis
    corresponding: false
bibliography: references.bib
csl: apa.csl
editor: 
  markdown: 
    wrap: 72
excecute:
  echo: true
---

# Introduction

Through this study, we hope to address and provide increased analysis of
the problem of firearm mortality in the United States. Gun violence is a
major safety, social, and public health problem within the United States
and requires additional investigation. The current literature suggests
that within developed nations and other high income countries, the U.S.
is an outlier in its rates of gun mortality. Initially linked to
America’s frontier history and culture of citizen-owned guns, the U.S.
continues to protect this 2nd amendment right and tradition of freedom,
protection, and spirit of rebellion (@fortune2018guns). While a
constitutional right, it may also be important to consider the needs of
present-day society and how this right can be protected, yet regulated
for a safer future. The large proportion of gun violence, mass
shootings, and firearm mortality can partially be attributed to the
weapon instrumentality effect. This is the hypothesis that the presence
of guns leads to higher levels of violence, aggression, and lethal
action when faced with a stressful or conflictual situation
(@spieller2020gun). While some argue that aggression and violence will
occur regardless of the weapon used, the capacity for lethality also has
to be considered. It is much easier to unintentionally harm someone with
a firearm than a knife, for instance. Understanding the relationship
between firearm ownership and firearm mortality through this study will
contribute to this system of thinking. If the hypothesis that firearm
ownership and firearm mortality are positively correlated, this study
can serve as further evidence that additional firearm restrictions
should be implemented to protect everyday American people. 

Within the U.S., gun mortality continues to be a major problem for its
population, contributing to both mortality and morbidity. In 2020 alone,
the Centers for Disease Control reported 45,000 firearm-related deaths
with a rate of 6.1 deaths per 100,000 people. This is the highest
reported rate since 1994 with 40% of deaths due to homicide and more
than half due to suicide (@fontanarosa2022unrelenting). Access to
firearms is also shifting suicide rates and the proportion of suicide
deaths by firearm. Within the 45,222 firearm deaths in 2020, 60% of
those were due to suicide. Similarly, among the 45,979 suicide deaths in
the same year, 51% of those were from a firearm injury
(@fontanarosa2022unrelenting). These statistics present a clear need for
additional firearm regulation. Our study hopes to provide further
evidence by illustrating a correlation between firearm ownership and
firearm mortality from the Centers for Disease Control and Prevention
and the World Population Review. 

# Methods

Two datasets were used in the research project, which were obtained from
the Centers for Disease Control and the World Population Review. The
data sets were semi-cleaned beforehand in Excel. Both data sets were
manipulated to include a year (2022) and state column with matching
variable names (AK changed to Alaska, etc). Information not needed for
analysis, like the URLs column in the gun mortality data set, was
removed for clarity. Once in R Studio, a full join was used to combine
the data sets for analysis. A final cleaning took place in R, where
renaming tools were used to make column category names clear and
concise. Filtering tools were used to ensure that only the 50 states
were included in the dataset, such as the District of Columbia, which is
already in the dataset from Washington. From the full join, some data
didn’t carry over, which resulted in the manual importing of data points
into the dataframe. Finally, the data was arranged based on the highest
to lowest gun ownership per household percentage.

First, a Shapiro test was conducted to test for normality in the data.
Once the data was determined to be normally distributed, a Pearson test
was used to look at the correlation between household gun ownership
percentage and mortality rate. Then, a linear regression model was used
to compare the top five states with the highest gun ownership to their
mortality rates. Finally, another linear regression model was used to
compare the top five states to the rest of the dataset to look for
correlations and assumptions.

# Results

```{r}
library(readr)
library(flextable)
library(dplyr)
library(ggplot2)
library(tidymodels)
library(ggpubr)

gun_mortality_2022 <- read.csv("data/gun_mortality_2022.csv")
gun_ownership_2022 <- read.csv("data/gun_ownership_2022.csv")

gun_data <- full_join(gun_mortality_2022, gun_ownership_2022, by = 'STATE')
gun_data <- gun_data |>
  rename(c(Year = YEAR.x,
         State = STATE,
         Death_Rate = RATE, 
         Deaths = DEATHS,
         Household_pct_ownership = GunOwnership_PercentageOfHouseholdsThatOwnGuns_pct_2022,
         GunOwnership = GunOwnership_NumOfGunLicenses_num_2022))|>
        select(State,Death_Rate,Deaths,Household_pct_ownership,GunOwnership,Year)
gun_data <- gun_data |>
  filter (State != "District of Columbia")

# Importing all values that didn't transfer over from the join
gun_data[51,3] <- as.integer(797)
gun_data[51,2] <- as.numeric(19.8)
gun_data[52,3] <- as.integer(830)
gun_data[52,2] <- as.numeric(14.0)
gun_data[53,3] <- as.integer(1316)
gun_data[53,2] <- as.numeric(14.9)
gun_data[36,4] <- as.numeric(54.7)
gun_data[36,5] <- as.integer(103368)
gun_data[46,4] <- as.numeric(44.6)
gun_data[46,5] <- as.integer(423707)
gun_data[49,4] <- as.numeric(45.3)
gun_data[49,5] <- as.integer(95430)
gun_data[51,6] <- as.integer(2022)
gun_data[52,6] <- as.integer(2022)
gun_data[53,6] <- as.integer(2022)

gun_data <- gun_data|>
  arrange(-Household_pct_ownership)

gun_data_con <- gun_data |>
  arrange(-Household_pct_ownership) |>
  head(n = 5)
```

Data Visualization tools were used to give a general idea of the
shortened dataset. Using the barplot function, a barplot was made with
the state as the x variable, and gun ownership % as the y variable.
Barplot color was based on death rates using the fill function, where
lower death rates were a darker blue, and higher death rates were a
lighter blue. Using barplots allowed the visualization data distribution
before the utilization of the Shapiro-Wilk test.

```{r}
plot1<- gun_data_con |>
ggbarplot(x = 'State', y = 'Household_pct_ownership',
          fill = "Death_Rate") 
print(plot1)
```

**Figure 1: Household gun ownership for top 5 states colored by death
rate.**

Multiple results were determined from the analysis of the shortened
dataset. It was determined through Shapiro-Wilk tests that both gun
ownership percentage & death rate had a p value greater than 0.05, with
a p value of 0.2029 for gun ownerhsip percentage among households, and a
p value of 0.5634 for death rates.

```{r}
shapiro.test(gun_data_con$Household_pct_ownership)

shapiro.test(gun_data_con$Death_Rate)

```

**Figure 2: Results of the Shapiro-Wilk normality test using the Pearson
method.**

Our Pearson method correlation test showed that the variables
Household_pct_ownership and Death_Rate had a correlation value of 0.905.
The Pearson correlation method was used because the sample data followed
a normal distribution. Plotting of Household_pct_ownership on the x-axis
and Death_Rate on the y-axis using a ggplot function allowed for
visualization of this relationship. A linear regression using death_rate
as our predicted variable, and household percent as our predictor
variable, was used to determine p value and r squared values to
determine if a linear relationship existed. With our sample data being
the data used, the model provided a p value of 0.03455, and an R-squared
value of 0.7591 

```{r}
correlation_metric <- cor(gun_data_con$Death_Rate, gun_data_con$Household_pct_ownership,  method = "pearson")
print(correlation_metric)

linear_mod <- lm(Death_Rate ~ Household_pct_ownership, data = gun_data_con)
summary(linear_mod)

```

**Figure 3: Summary results on linear regression model using gun
ownership as a predictor for death rate.**

```{r}
plot2 <- ggplot(gun_data_con, aes(x = Household_pct_ownership, y = Death_Rate)) +
  geom_point() +
  geom_smooth(method = lm, color = 'red') +
  labs(x = "Gun % Per Household",
       y = "Death Rate",
       title = "Relationship Between gun % Per Household and Death Rate ") +
  theme_linedraw()
print(plot2)
```

**Figure 4: Relationship between death rate and gun ownership % per
household.**

A predicted graph was made to visualize the relationship between gun
ownership per household and death rate to determine the accuracy of our
model. Using lm(household_pct_ownership \~ death_rate, data =
gun_data_con) representing the cleaned dataset with the top 5 states for
gun ownership, this fit was applied to the full cleaned dataset and was
then visualized. This visualization found that datapoints that had lower
gun ownership percentages had higher residual values than those with
higher gun ownership %.

```{r}
rec <- recipe(Household_pct_ownership ~ Death_Rate, data = gun_data_con)
prep <- prep(rec)
baked_data <- bake(prep, new_data = gun_data_con)

lm_pred <- lm(Household_pct_ownership ~ Death_Rate, data = baked_data)
summary(lm_pred)

prediction_data <- augment(lm_pred, new_data = gun_data)

```

```{r}
plot4 <- gun_data %>%
ggplot(aes(x = Household_pct_ownership, y = Death_Rate)) +
  geom_point(col = "blue") +
  geom_smooth(method = "lm", col = "red") + 
  labs(xlab = "Gun Ownership Percentage", ylab = "Observed Death Rate", 
         title = "Observed Death Rate Based on Gun Ownership Percentage")
print(plot4)
```

**Figure 5: Observed death rate based on gun ownership % in all
states.**

```{r}
pred_data <- augment(linear_mod, newdata = gun_data)

plot5 <- pred_data %>%
ggplot(aes(x = Household_pct_ownership, y = .fitted)) +
  geom_point(aes(color = .resid))
print(plot5)
```

**Figure 6: Predicted death rate values using household ownership %
using linear regression from the top 5 states colored by residuals.**

# Discussion

After completing our analysis, the results show a strong correlation
between gun ownership and gun mortality when focused specifically on the
five states with the highest levels of ownership. As displayed in Figure
1, the top five states for gun ownership were Montana, Wyoming, Alaska,
Idaho, and West Virginia. Based on the Pearson test, the two variables,
firearm ownership and firearm mortality, show a high correlation. The
resulting correlation value of 0.905 is very close to 1, showing a
strong positive correlation between the two variables. Given the results
of this test, it made sense to complete a linear regression. From the
linear regression, a p-value of 0.03455 was produced. This shows
statistical significance because the value is below 0.05. Additionally,
the resulting r-squared value of 0.7591 is between the range of
0.7-0.99, showing significance. In essence, as displayed in Figure 3,
our results show that there is a strong positive correlation between
firearm ownership and firearm mortality. Owning a firearm is a direct
contributing factor to deaths caused by firearms, at least at the state
level. 

In terms of the broader implications of this study, these results may be
used as evidence for further reform to firearm ownership requirements in
the United States. More restrictions on the number of firearms one
person may own could help reduce the number of firearm-related deaths
seen in these areas. Additionally, enacting common-sense gun laws that
require additional screenings for gun purchasing and the types of
firearms available to the public may be necessary. For instance, placing
restrictions on the type of firearms available for public purchase could
limit the number of deaths resulting from automatic rifles per region.
If the states with fewer firearms have fewer deaths caused by firearms,
it makes sense to place limits and additional restrictions on the number
of firearms one may possess. 

In terms of areas for improvement, this study contains a few
limitations. One limitation would be the scope of this project. For
simplicity's sake, our project focused on the top five states with the
highest levels of gun ownership. However, given that there are 50 states
within the United States, this means only 10 percent of the total states
were included in the analysis. In future analysis, more, if not all, of
the states could be included to gain a more complete understanding of
the correlation present. Alternatively, in addition to the top five
states for ownership, this study could have included the bottom 5 states
for ownership to provide a comparison at the other end of the spectrum.
Additionally, our study is limited in terms of supplemental data sets.
While this study provides a clear correlation between gun ownership and
gun mortality, it is lacking in additional variables and contributing
factors. For instance, population density could be a contributing factor
to this relationship. Do areas with a higher population density have
more or less firearm mortality as a result? This might be an important
topic to investigate in further studies.  

Furthermore, this study relates to the broader field of data science by
showing how data science can be employed as a tool to inform social
issues. This type of study is important as it takes raw data and
transforms it into something tangible for the public and government
leaders to understand. Rather than developing policy decisions purely
based on ideas of what might be going on, this study supports reformed
gun laws through concrete evidence. This is also a demonstration of how
data science can be used to model and address complex, real-world
problems such as gun mortality in the United States. Additionally,
studies such as this one contribute to clearer communication of results
within the scientific community. Being able to produce compelling
charts, graphs, and other visualizations makes data easier to digest for
a lay audience. For instance, in this study, Figure 4 displays the
linear regression relating firearm ownership and firearm mortality,
providing a clear representation of this relationship. While the average
American may not have knowledge of a Pearson test or even be familiar
with the verbiage of a linear regression, they likely have a basic
understanding of graphs and can identify a linear relationship. Making
studies like this available to the public plays a key role in
contributing to informed policy decisions and bridging the gap between
data and the real-world implications. 

As for next steps, replication of the Pearson correlation test and
linear regression using more states would be ideal. Addressing one of
the study’s limitations, this would provide even greater evidence for
the correlation identified through this study. At this point, the
correlation identified in this study could be used as evidence for
reform to the current firearm ownership process. Encouraging
restrictions on how many firearms may be purchased by a single
individual or requiring additional screenings would be one approach. 

For even greater and potentially more specific evidence, additional
variables may be added to this analysis. As aforementioned, population
density by state could be an important consideration. If taking this
approach, narrowing the region of focus to the county level may also be
advantageous. Moreover, data on firearm type could also be studied. Data
identifying gun ownership by type of gun (e.g., handgun, hunting rifle,
shotgun), as well as semi-automatic and automatic, may be used to
provide more specific insight into the firearm regulation needed. For
example, if findings illustrate that areas with a higher number of
automatic rifles also have the highest number of firearm mortality, this
may be used as evidence for placing further restrictions on automatic
rifle ownership. If using firearm type as a contributing variable, it
may be necessary to also find more specific data on firearm mortality
rates. Specifically, within the firearm mortality data, identifying what
type of firearm that was responsible for the mortality. With these
changes, it may be possible to identify which firearms contribute most
to firearm mortality and if there is a correlation between ownership of
these types of firearms and mortality caused by them. 

Overall, the statistical tests used and results presented in this study
demonstrate a clear positive correlation between firearm ownership and
firearm mortality within the top five states for firearm ownership. This
study contributes to the growing evidence that further restrictions on
firearm ownership are necessary and also serves as a starting point for
investigations into other contributing factors of firearm mortality.

# References
