---
title: Comparing Firearm Mortality to Firearm Ownership by State
authors:
  - name: Grace Rinehart
    affiliation: Colorado State University
    roles: writing
    corresponding: true
  - name: Kendall Landwehr
    affiliation: Colorado State University
    roles: research
    corresponding: false
  - name: Andrew Zimbelman
    affiliation: Colorado State University
    roles: data collection & analysis
    corresponding: false
bibliography: references.bib
csl: apa.csl
editor: 
  markdown: 
    wrap: 72
excecute:
  echo: true
---

# Abstract

There is a notion that having more firearms makes a country or state
“safer.” However, having more firearms per capita has been proven to
worsen mortality rates. This issue is especially prevalent in the United
States, where 2nd amendment rights have become a topic of debate. In
this study, we aim to provide more evidence that states with high rates
of firearm ownership have higher rates of firearm mortality. Using
several statistical analyses, we highlight the correlations between
firearm ownership and firearm mortality. Our results showed that there
is a significant, positive correlation between firearm ownership and
firearm mortality. The states with the highest rates of firearm
ownership equally had the highest rates of firearm mortality. This trend
was demonstrated to be exponential across the five states with the
highest firearm ownership percentages, as shown in both a linear
regression model and a bar plot. By providing more insight into firearm
mortality in relation to firearm ownership, this study hopes to
encourage more rigorous laws regarding firearm safety and ownership
requirements.

# Introduction

Through this study, we hope to address and provide increased analysis of
the problem of firearm mortality in the United States. Gun violence is a
major safety, social, and public health problem within the United States
and requires additional investigation. America has the highest
percentage of gun ownership with an estimated 270 to 310 million
privately owned guns in the country (@stroebe2017dangerous). This is
drastically higher than any other country. For instance, Germany has
about 15.8 million privately owned firearms, Canada and France both have
12.7 million, Brazil has 17.8 million, and South Africa has 5.4 million
(@karp2018estimating). In a 2024 study with a demographically diverse
survey of gun owners, the majority reported that they purchased a gun
for personal protection and safety (@ward2024reasons). With the goal of
protection in mind, the high volume of firearm death in the United
States is concerning. While these firearms are being purchased with the
intention to protect, they may be causing more harm than good. 

Gun violence has been a problem for decades, but there was a rapid
increase following the COVID-19 pandemic. Homicides alone increased by
30% from 2019 to 2020 and there was an additional 7% increase in 2021
(@braga2022gun). This shows the continued relevance of this topic and
how developing solutions to this crisis remains necessary. The current
literature suggests that within developed nations and other high income
countries, the U.S. is an outlier in its rates of gun mortality.
Initially linked to America’s frontier history and culture of
citizen-owned guns, the U.S. continues to protect this 2nd amendment
right and tradition of freedom, protection, and spirit of rebellion
(@fortune2018guns). While a constitutional right, it may also be
important to consider the needs of present-day society and how this
right can be protected, yet regulated for a safer future. The large
proportion of gun violence, mass shootings, and firearm mortality can
partially be attributed to the weapon instrumentality effect. This is
the hypothesis that the presence of guns leads to higher levels of
violence, aggression, and lethal action when faced with a stressful or
conflictual situation (@spieller2020gun). While some argue that
aggression and violence will occur regardless of the weapon used, the
capacity for lethality also has to be considered. It is much easier to
unintentionally harm or kill someone with a firearm than a knife, for
instance. Understanding the relationship between firearm ownership and
firearm mortality through this study will contribute to this system of
thinking. If the hypothesis that firearm ownership and firearm mortality
are positively correlated, this study can serve as further evidence that
additional firearm restrictions should be implemented to protect
everyday American people.

Within the U.S., gun mortality continues to be a major problem for its
population, contributing to both mortality and morbidity. In 2020 alone,
the Centers for Disease Control reported 45,000 firearm-related deaths
with a rate of 6.1 deaths per 100,000 people. This is the highest
reported rate since 1994 with 40% of deaths due to homicide and more
than half due to suicide (@fontanarosa2022unrelenting). Access to
firearms is also shifting suicide rates and the proportion of suicide
deaths by firearm. Within the 45,222 firearm deaths in 2020, 60% of
those were due to suicide. Similarly, among the 45,979 suicide deaths in
the same year, 51% of those were from a firearm injury
(@fontanarosa2022unrelenting). These statistics present a clear need for
additional firearm regulation. Our study hopes to provide further
evidence by illustrating a correlation between firearm ownership and
firearm mortality from the Centers for Disease Control and Prevention
and the World Population Review.

# Methods

Two datasets were used in the research project, which were obtained from
the Centers for Disease Control and the World Population Review. The
data sets were semi-cleaned beforehand in Excel. Both data sets were
manipulated to include a year (2022) and state column with matching
variable names (AK changed to Alaska, etc). Information not needed for
analysis, like the URLs column in the gun mortality data set, was
removed for clarity. Once in R Studio, a full join was used to combine
the data sets for analysis. A final cleaning took place in R, where
renaming tools were used to make column category names clear and
concise. Filtering tools were used to ensure that only the 50 states
were included in the dataset, such as the District of Columbia, which is
already in the dataset from Washington. From the full join, some data
didn’t carry over, which resulted in the manual importing of data points
into the dataframe. Finally, the data was arranged based on the highest
to lowest gun ownership per household percentage.

First, a Shapiro test was conducted to test for normality in the data.
Once the data was determined to be normally distributed, a Pearson test
was used to look at the correlation between household gun ownership
percentage and mortality rate. Then, a linear regression model was used
to compare the top five states with the highest gun ownership to their
mortality rates. Finally, another linear regression model was used to
compare the top five states to the rest of the dataset to look for
correlations and assumptions.

# Results

```{r}
library(readr)
library(flextable)
library(dplyr)
library(ggplot2)
library(tidymodels)
library(ggpubr)

gun_mortality_2022 <- read.csv("data/gun_mortality_2022.csv")
gun_ownership_2022 <- read.csv("data/gun_ownership_2022.csv")

gun_data <- full_join(gun_mortality_2022, gun_ownership_2022, by = 'STATE')
gun_data <- gun_data |>
  rename(c(Year = YEAR.x,
         State = STATE,
         Death_Rate = RATE, 
         Deaths = DEATHS,
         Household_pct_ownership = GunOwnership_PercentageOfHouseholdsThatOwnGuns_pct_2022,
         GunOwnership = GunOwnership_NumOfGunLicenses_num_2022))|>
        select(State,Death_Rate,Deaths,Household_pct_ownership,GunOwnership,Year)
gun_data <- gun_data |>
  filter (State != "District of Columbia")

# Importing all values that didn't transfer over from the join
gun_data[51,3] <- as.integer(797)
gun_data[51,2] <- as.numeric(19.8)
gun_data[52,3] <- as.integer(830)
gun_data[52,2] <- as.numeric(14.0)
gun_data[53,3] <- as.integer(1316)
gun_data[53,2] <- as.numeric(14.9)
gun_data[36,4] <- as.numeric(54.7)
gun_data[36,5] <- as.integer(103368)
gun_data[46,4] <- as.numeric(44.6)
gun_data[46,5] <- as.integer(423707)
gun_data[49,4] <- as.numeric(45.3)
gun_data[49,5] <- as.integer(95430)
gun_data[51,6] <- as.integer(2022)
gun_data[52,6] <- as.integer(2022)
gun_data[53,6] <- as.integer(2022)

gun_data <- gun_data|>
  arrange(-Household_pct_ownership)

gun_data_con <- gun_data |>
  arrange(-Household_pct_ownership) |>
  head(n = 5)
```

Data Visualization tools were used to give a general idea of the
shortened dataset. Using the barplot function, a barplot was made with
the state as the x variable, and gun ownership % as the y variable.
Barplot color was based on death rates using the fill function, where
lower death rates were a darker blue, and higher death rates were a
lighter blue. Using barplots allowed the visualization data distribution
before the utilization of the Shapiro-Wilk test.

```{r}
plot1<- gun_data_con |>
ggbarplot(x = 'State', y = 'Household_pct_ownership',
          fill = "Death_Rate") 
print(plot1)
```

**Figure 1: Barplot visualizing the household ownership based on state,
colored by death rate. This shows the relationship between household gun
ownership percentage and death rate.**

Multiple results were determined from the analysis of the shortened
dataset. It was determined through Shapiro-Wilk tests that both gun
ownership percentage & death rate had a p value greater than 0.05, with
a p value of 0.2029 for gun ownerhsip percentage among households, and a
p value of 0.5634 for death rates.

```{r}
shapiro.test(gun_data_con$Household_pct_ownership)

shapiro.test(gun_data_con$Death_Rate)

```

**Figure 2: Results of the Shapiro-Wilk normality test using the Pearson
method. The test showed a normal distribution.**

Our Pearson method correlation test showed that the variables
Household_pct_ownership and Death_Rate had a correlation value of 0.905.
The Pearson correlation method was used because the sample data followed
a normal distribution. Plotting of Household_pct_ownership on the x-axis
and Death_Rate on the y-axis using a ggplot function allowed for
visualization of this relationship. A linear regression using death_rate
as our predicted variable, and household percent as our predictor
variable, was used to determine p value and r squared values to
determine if a linear relationship existed. With our sample data being
the data used, the model provided a p value of 0.03455, and an R-squared
value of 0.7591 

```{r}
correlation_metric <- cor(gun_data_con$Death_Rate, gun_data_con$Household_pct_ownership,  method = "pearson")
print(correlation_metric)

linear_mod <- lm(Death_Rate ~ Household_pct_ownership, data = gun_data_con)
summary(linear_mod)

```

**Figure 3a: Summary results on linear regression model using gun
ownership as a predictor for death rate. The regression shows a strong
relationship between household gun ownership percentage and death
rate.**

```{r}
plot2 <- ggplot(gun_data_con, aes(x = Household_pct_ownership, y = Death_Rate)) +
  geom_point() +
  geom_smooth(method = lm, color = 'red') +
  labs(x = "Gun % Per Household",
       y = "Death Rate",
       title = "Relationship Between gun % Per Household and Death Rate ") +
  theme_linedraw()
print(plot2)
```

**Figure 3b:** **Graph visualizing the relationship between household
gun ownership percentage and death rate. This shows that both variables
linearly increase together.**

A predicted graph was made to visualize the relationship between gun
ownership per household and death rate to determine the accuracy of our
model. Using lm(household_pct_ownership \~ death_rate, data =
gun_data_con) representing the cleaned dataset with the top 5 states for
gun ownership, this fit was applied to the full cleaned dataset and was
then visualized. This visualization found that datapoints that had lower
gun ownership percentages had higher residual values than those with
higher gun ownership percentages.

```{r}
rec <- recipe(Household_pct_ownership ~ Death_Rate, data = gun_data_con)
prep <- prep(rec)
baked_data <- bake(prep, new_data = gun_data_con)

lm_pred <- lm(Household_pct_ownership ~ Death_Rate, data = baked_data)
summary(lm_pred)

prediction_data <- augment(lm_pred, new_data = gun_data)

```

**Figure 4a: Summary results on linear regression model using death rate
as a predictor for gun ownership. The regression shows a strong
relationship between death rate and household gun ownership
percentage.**

```{r}
pred_data <- augment(linear_mod, newdata = gun_data)

plot5 <- pred_data %>%
ggplot(aes(x = Household_pct_ownership, y = .fitted)) +
  geom_point(aes(color = .resid))
print(plot5)
```

**Figure 4b: Graph visualizing predicted death rate values using
household ownership percent with a linear regression from the top five
states, colored by residuals. This graph shows that residual values
increase as data moves away from our sample data.**

# Discussion

After completing our analysis, the results show a strong correlation
between gun ownership and gun mortality when focused specifically on the
five states with the highest levels of ownership. As displayed in Figure
1, the top five states for gun ownership were Montana, Wyoming, Alaska,
Idaho, and West Virginia. Based on the Pearson test, the two variables,
firearm ownership and firearm mortality, show a direct correlation. The
resulting correlation value of 0.905 is very close to 1, showing a
strong positive correlation between the two variables (Figure 2). Given
the results of this test, it made sense to complete a linear regression.
From the linear regression, a p-value of 0.03455 was produced (Figure
3a). This shows statistical significance because the value is below
0.05. Additionally, the resulting r-squared value of 0.7591 is between
the range of 0.7-0.99, showing significance (Figure 3a). Examining the
relationship between the percentage of households that own a firearm and
firearm mortality, there is a positive direct correlation. As the
percentage of households that own a firearm increases, so does the
firearm mortality rate (Figure 3b). In essence, as displayed in Figure
4b, our results show that there is a strong positive correlation between
firearm ownership and firearm mortality. Owning a firearm is a direct
contributing factor to deaths caused by firearms, at least at the state
level. 

In terms of the broader implications of this study, these results may be
used as evidence for further reform to firearm ownership requirements in
the United States. More restrictions on the number of firearms one
person may own could help reduce the number of firearm-related deaths
seen in these areas. Additionally, enacting common-sense gun laws that
require additional screenings for gun purchasing and the types of
firearms available to the public may be necessary. For instance,
existing literature exploring the relationship between state firearm
restrictions and suicide rates shows that states with background checks,
mandatory waiting periods, and prohibited possession show lower suicide
rates than states without these same regulations
(@kawano2023restrictive). If the states with fewer and more regulated
firearms have fewer deaths caused by firearms, it makes sense to place
limits and additional restrictions on the number of firearms one may
possess. 

In terms of areas for improvement, this study contains a few
limitations. One limitation would be the scope of this project. For
simplicity's sake, our project focused on the top five states with the
highest levels of gun ownership. However, given that there are 50 states
within the United States, this means only 10 percent of the total states
were included in the analysis. In future analysis, more, if not all, of
the states could be included to fully understand the correlation
present. Alternatively, in addition to the top five states for
ownership, this study could have included the bottom 5 states for
ownership to provide a comparison at the other end of the spectrum.
Additionally, our study is limited in terms of supplemental data sets.
While this study provides a clear correlation between gun ownership and
gun mortality, it is lacking in additional variables and contributing
factors. As described in an article by the American Psychological
Association, many risk factors may increase one’s likelihood of
participating in gun violence, resulting in firearm mortality. These
factors include lack of employment, socioeconomic status, safety
concerns, and community distress (@wamser2021understanding). Finding
relevant and corresponding data sets could help extend the analysis in
this study and quantify the impact of these specific risk factors on
firearm mortality. 

Furthermore, this study relates to the broader field of data science by
showing how data science can be employed as a tool to inform social
issues. This type of study is important as it takes raw data and
transforms it into something tangible for the public and government
leaders to understand. As illustrated in a ScienceDirect article, data
science is crucial in helping make informed policy decisions in
government, from the municipal to the federal level. Just as data
science can provide insight and key visualizations into a city’s
transportation sector, it can also be used to inform decisions regarding
firearm regulations (@matheus2020data). Rather than developing policy
decisions purely based on ideas of what might be going on, this study
supports reformed gun laws through concrete evidence. This is also a
demonstration of how data science can be used to model and address
complex, real-world problems such as gun mortality in the United States.
Studies such as this also contribute to clearer communication of results
within the scientific community. Being able to produce compelling
charts, graphs, and visualizations makes data easier to digest for a lay
audience. For instance, in this study, Figures 4 and 5 display the
linear regression relating firearm ownership and mortality, clearly
representing this relationship. While the average American may not have
knowledge of a Pearson test or be familiar with the verbiage of a linear
regression, they likely have a basic understanding of graphs and can
identify a linear relationship. Making studies like this available to
the public is key in contributing to informed policy decisions and
bridging the gap between data and real-world implications.

As for next steps, replication of the Pearson correlation test and
linear regression using more states would be ideal. Addressing one of
the study’s limitations, this would provide even greater evidence for
the correlation identified through this study. At this point, the
correlation identified in this study could be used as evidence for
reform to the current firearm ownership process. Encouraging
restrictions on how many firearms may be purchased by a single
individual or requiring additional screenings would be one approach. 

Additional variables and risk factors may be added to this analysis for
even greater and potentially more specific evidence. If taking this
approach, narrowing the region of focus to the county level may also be
advantageous. Moreover, data on firearm type could also be studied. Data
identifying gun ownership by type of gun (e.g., handgun, hunting rifle,
shotgun), semi-automatic, and automatic, may be used to provide more
specific insight into the firearm regulation needed. For example, if
findings illustrate that areas with a higher number of automatic rifles
also have the highest number of firearm mortality, this may be used as
evidence for placing further restrictions on automatic rifle ownership.
If using firearm type as a contributing variable, it may be necessary to
also find more specific data on firearm mortality rates. Specifically,
within the firearm mortality data, identifying what type of firearm that
was responsible for the mortality. With these changes, it may be
possible to identify which firearms contribute most to firearm mortality
and if there is a correlation between ownership of these types of
firearms and mortality caused by them.

# Conclusion

Overall, the Pearson test performed in this analysis resulted in a
statistically significant R-squared and p-value. Likewise, linear
regression demonstrates a clear positive correlation between firearm
ownership and firearm mortality within the top five states for firearm
ownership. This study contributes to the growing evidence that further
restrictions on firearm ownership are necessary. Evidence suggests that
with a smaller percentage of households owning firearms, it is likely
that firearm mortality will decrease. Furthermore, this study serves as
a starting point for investigations into other contributing factors of
firearm mortality, in order to make specific and impactful firearm
restrictions.  

# References
